\section{Phased Computation Graphs}
\label{sec:pcg}

\begin{table}[t]
\small
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|} \hline
Model of Computation & Items on Channels & Nodes with & Nodes with & Cyclic Steady & Initialization \\
                     & at Start & Pop $\ne$ Push & Peek $>$ Pop & State Phases & Phases \\
\hline \hline
Synchronous Dataflow~\cite{LM87-i} & X & X & & & \\
\hline
Cyclo-Static Dataflow~\cite{BELP96} & X & X & & X & \\
\hline
Computation Graphs~\cite{KM66} & X & X & X & & \\
\hline
Phased Computation Graphs & X & X & X & X & X \\
\hline
\end{tabular}
\vspace{-6pt}
\caption{\protect\small Models of Computation for Dataflow Graphs.}
\label{tab:models}
\vspace{-12pt}
\end{center}
\end{table}

We introduce the Phased Computation Graph (PCG) as a model of
computation for dataflow languages; specifically, we designed the PCG
as a model for StreamIt~\cite{streamitcc}.  A PCG is a generalization
of the Computation Graph (CG) of Karp and Miller~\cite{KM66} to the
case where each node has both an initial and steady-state execution
epoch, each of which contains a number of {\it phases}.  It is also a
generalization of other popular representations for dataflow graphs,
such as synchronous dataflow (SDF) and cyclo-static dataflow (CSDF)
(see Table~\ref{tab:models}).  Phases are attractive constructs for
the reasons given in describing CSDF in Section~\ref{sec:related};
also, an initial epoch is critical for applications such as a WAV file
decoder that adjusts its parameters based on the data it finds in the
header of the stream.

Formally, a PCG is a directed graph with the following components:
\begin{itemize}

\item Nodes $n_1 \dots n_{m_n}$.

\item Channels $c_1 \dots c_{m_c}$, where each channel is directed
from a given node $n_a$ to a given node $n_b$.

\item A non-negative integer $A(c)$ indicating the number of items
  that are initially present on channel $c$.

\item The number of phases $\mt{num}(n, t)$ that node $n$ exhibits
during epoch $t$.  Here $t$ indicates the type of epoch, which is
either initial ($\mt{init}$) or steady-state ($\mt{steady}$).  For all
$n$, $\mt{num}(n, \mt{init}) \ge 0$ and $\mt{num}(n, \mt{steady}) \ge
1$.

\item Non-negative integers $U(c, t, p)$, $O(c, t, p)$, and $E(c, t,
p)$, which give the number of items pushed, popped, and peeked,
respectively, over channel $c$ during the $p$'th phase of epoch $t$.

\end{itemize}

\subsection{Execution Model}

We give an informal description of the execution semantics of a PCG,
which is an intuitive extension to that of a CG.  An execution
consists of a sequence of atomic {\it firings} of nodes in the graph.
The effect of firing a given node depends on its epoch and its phase,
which are both local states of the node.  At the start of execution,
each node is in phase 0 of the initial epoch, and each channel $c$
contains $A(c)$ items.

Let us consider a node $n$ that is in phase $p$ of epoch $t$.  It is
legal for $n$ to fire if $p \in [0, \mt{num}(n, t)-1]$ and, for each
channel $c_{in}$ directed into $n$, there are at least $E(c_{in}, t,
p)$ items on $c_{in}$.  The effect of this firing is to consume
$O(c_{in}, t, p)$ items from each channel $c_{in}$ directed into $n$;
to produce $U(c_{out}, t, p)$ new items on each channel $c_{out}$
directed out of $n$; and to advance the phase $p$ of $n$ to
$\mt{next}(t, p)$.  In the initial epoch, each phase executes only
once and $\mt{next}(\mt{init}, p) = p + 1$; in the steady-state epoch,
the phases execute cyclically and $\mt{next}(\mt{steady}, p) = (p +
1)~\mt{mod}~\mt{num}(n, t)$.  If a node $n$ is in phase $\mt{num}(n,
\mt{init})$ of the initial epoch, then it transitions to the
steady-state epoch and resets its phase to 0.

From the starting state of the graph, execution proceeds via an
infinite sequence of node firings.  The order of node firings is
constrained only by the conditions given above.

\section{Phased Computation Programs}

The PCG described above is a representation for a graph, treating each
node and channel as a black box.  Here we introduce some notation for
the internals of the nodes, as well as the ordering of items on the
channels.  We refer to the aggregate model as a Phased Computation
Program (PCP).  Before describing a PCP, we will need some additional
notation:
\begin{itemize}

\item We assume that all items passed over channels are drawn from the
same domain of values ${\cal V}$.

\item An array type\footnote{To simplify the presentation, we allow
ourselves a slight abuse of notation by sometimes using arrays instead
of enumerating individual elements.  Our definitions could be expanded
into strict element-wise SARE's without any fundamental change to the
technique.} of length $n$ and element type $\tau$ is denoted by
$\tau[n]$.  Given a two-dimensional array $A$ of type $\tau[n][m]$,
$A[i][*]$ denotes the $m$-element array comprising the $i$'th row of
$A$.

\item $\mt{num\_in}(n)$ (resp. $\mt{num\_out}(n)$) denotes the number of
channels that are directed into (resp. out of) node $n$.

\item $\mt{chan\_in}(n)$ (resp. $\mt{chan\_out}(n)$) denotes the list of
channels that are directed into (resp. out of) node $n$.

\item $\mt{pos\_in}(n, c)$ (resp. $\mt{pos\_out}(n, c)$) denotes the
position of channel $c$ in $\mt{chan\_in}(n)$
(resp. $\mt{chan\_out}(n)$.  That is, $\mt{chan\_out}(n)[\mt{pos\_out(n,
c)}] = c$.

\end{itemize}

\noindent A PCP consists of the following:
\begin{itemize}

\item A Phased Computation Graph $G = (\{n_1 \dots n_{m_n}\}, \{c_1
\dots c_{m_c}\}, A, \mt{num}, U, O, E)$ describing the nodes,
channels, and I/O rates of the program (see Section~\ref{sec:pcg}).
Each channel $c = (n_a, n_b)$ is a FIFO queue; $n_a$ pushes items onto
the back of $c$ and $n_b$ consumes items from the front of $c$.

\item The initial values $I(c)$ that are enqueued onto channel $c$ at
the start of execution.  Since there are $A(c)$ initial values on
channel $c$, $I(c)$ has type ${\cal V}[A(c)]$.

%An initialization function $I(c)$ that returns an array of
%length $A(c)$, the elements of which are enqueued onto channel $c$ at
%the start of execution.  The procedure $I(c)$ takes no arguments; its
%signature is $\mt{void} {\small \rightarrow} {\cal V}[A(c)]$.
%
\item A work function $W(n, t, p)$ that represents the computation of
node $n$ in phase $p$ and epoch $t$.  The signature of $W(n, t, p)$ is
$[{\cal V}[E(\mt{chan\_in}(n)[1], t, p)], \dots, {\cal
V}[E(\mt{chan\_in}(n)[\mt{num\_in}(n)], t, p)]]$ $\rightarrow$ \\ $[
{\cal V}[U(\mt{chan\_out}(n)[1], t, p)], \dots, {\cal
V}[U(\mt{chan\_out}(n)[\mt{num\_out}(n)], t, p)]]$.  That is, the
function inputs a list of arrays, each of which corresponds to an
incoming channel $c_{in}$ and contains the $E(c_{in},t,p)$ values that
$n$ reads from $c_{in}$ in a given firing.  The procedure returns a
list of arrays, each of which corresponds to an outgoing channel
$c_{out}$ and contains the $U(c_{out},t,p)$ values that $n$ writes to
$c_{out}$ in a given firing.

\end{itemize}

%The components $I$ and $W$ above are referred to as ``procedures''
%rather than mathematical functions because we expect them to be
%represented as blocks of source code for an actual procedure
%declaration, thereby allowing a straightforward conversion to a PCP
%from a dataflow program.  However, 

