\section{Related Work}
\label{sec:related}

Several other researchers have pursued the idea of operating directly
on compressed data formats.  The novelty of our work is two-fold:
first, in its ability to map an arbitrary stream program, rather than
a single predefined operation, into the compressed domain; and second,
in its focus on lossless compression formats.

Most of the previous work on mapping algorithms into the compressed
domain has focussed on formats such as JPEG that utilize a Discrete
Cosine Transform (DCT) to achieve spatial
compression~\cite{refs-from-intro}.  This task requires a different
analysis, with particular attention given to details such as the
blocked decomposition of the image, quantization of DCT coefficients,
zig-zag ordering, and so-on.  Because there is also a run-length
encoding stage in JPEG, our current technique might find some
application there; however, it appears that techniques designed for
JPEG have limited application to formats such as LZ77.  Also, we are
unaware of any previous methodology for translating a generic program
to operate on compressed data; previous efforts have mapped each
algorithm in a manual and ad-hoc way.

There has been some interest in performing compressed processing on
lossless encodings of black-and-white images.  Shoji presents the pxy
format for performing transpose and other affine
operations~\cite{shoji95}; the memory behavior of the technique was
later improved by Misra et al.~\cite{images/99-misra}.  As described
in Section~\ref{sec:formats}, the pxy format lists the $(x,y)$
coordinate pairs at which a black-and-white image changes color during
a horizontal scan.  If the image is represented using LZ77
compression, our transformation could also preserve some compression
during a transpose if we include support for filters with state (see
Section~\ref{sec:extensions}.

From the perspective of programming languages, Swartz and Smith
present RIVL, a Resolution Independent Video
Language~\cite{languages/swartz95}.  The language is used to describe
a sequence of image transformations; this allows the compiler to
analyze the sequence and, via lazy evaluation, to eliminate any
operations that do not effect the final output.  Such a technique is
complementary to ours and could also be implemented using StreamIt as
the source language.

%% 11. B. Shen and I. K. Sethi, ``Direct feature extraction from
%% compressed images,’’ in Storage & Retrieval for Image and Video
%% Databases IV, Proc. SPIE 2670, pp. 404-414 ~1996.
%%
%% 12. B. Shen and I. K. Sethi, ``Convolution-based Edge detection
%% for image/video in block DCT domain,’’ J. Visual Commun. Image
%% Represent 7~4'', 411-423 ~1996!.
%%
%% Brian C. Smith and Lawrence A. Rowe. Compressed domain processing of
%% JPEG encoded images. Real-Time Imaging, 2, 1996.
