\section{Conclusions and Future Work}
\label{sec:conc}

This paper presents StreamIt, a novel language for high-performance
streaming applications.  Stream programs are emerging as a very
important class of applications with distinct properties from other
recognized application classes.  This paper develops fundamental
programming constructs for the streaming domain.

The primary goal of StreamIt is to raise the abstraction level in
stream programming without sacrificing performance.  We have argued
that StreamIt's mechanisms for filter definition, filter composition,
messaging, and re-initialization will improve programmer productivity
and program robustness within the streaming domain.

Also, we believe that StreamIt is a viable common machine language for
grid-based architectures (e.g., \cite{smartmemories,rawshort,trips}),
just as C is a common machine language for von-Neumann machines.
StreamIt abstracts away the target's granularity, memory layout, and
network interconnect, while capturing the notion of independent
processors that communicate in regular patterns.  Fission and fusion
algorithms can automatically adjust the granularity of a stream graph
to match that of a given target.

We have a number of extensions planned for the next version of the
StreamIt language.  The current version is designed primarily for
uniform one-dimensional data processing, but constructs for hierarchical
frames of data would be useful for image processing.  Moreover, a future
version will support dynamically varying I/O rates of the filters in the
stream.  We expect that such support will require new language
constructs--for instance, a type-dispatch splitter that routes items to
the components of a split-join based on their type, and a fall-through
joiner that pulls items from any stream in a split-join as soon as they
are produced.

Our  immediate focus  is on  developing a  high-performance optimizing
compiler  for StreamIt that  can match  the performance  of hand-coded
applications, such that the abstraction benefits of StreamIt come with
no performance penalty.

\section*{Acknowledgments}

% The StreamIt compiler was implemented with Michael Gordon and David
% Maze, with applications support of Jeremy Wong, Henry Hoffman, and
% Matthew Brown; we also thank Matt Frank for many helpful comments.
This work was supported in part by the MIT Oxygen Project and DARPA
Grant DBT6396-C-0036.
