\section{Related Work}
\label{sec:related}
A large number of programming languages have included a concept of a
stream; see \cite{survey97} for a survey.  Those that are perhaps most
related to StreamIt 1.0 are synchronous dataflow languages such as
LUSTRE~\cite{lustre} and ESTEREL~\cite{esterel92}, which require a fixed
number of inputs to arrive simultaneously before firing a stream node.
However, most special-purpose stream languages do not contain features
such as messaging and support for modular program development that are
essential for modern stream applications.  Also, most of these languages
are so abstract and unstructured that the compiler cannot perform enough
analysis and optimization to result in an efficient implementation.

At an abstract level, the stream graphs of StreamIt share a number of
properties with the synchronous dataflow (SDF) domain as considered by
the Ptolemy project~\cite{ptolemyoverview}.  Each node in an SDF graph
produces and consumes a given number of items, and there can be delays
along the arcs between nodes (corresponding loosely to items that are
peeked in StreamIt).  As in StreamIt, SDF graphs are guaranteed to
have a static schedule and there are a number of nice scheduling
results incorporating code size and execution time~\cite{leesdf}.
However, previous results on SDF scheduling do not consider
constraints imposed by point-to-point messages, and do not include a
notion of StreamIt's information wavefronts, re-initialization, and
programming language support.

A specification package used in industry bearing some likeness to
StreamIt is SDL: Specification and Description Language~\cite{sdlrec}.
SDL is a formal, object-oriented language for describing the structure
and behavior of large, real-time systems, especially for
telecommunications applications.  It includes a notion of asynchronous
messaging based on queues at the receiver, but does not incorporate
wavefront semantics as does StreamIt.  Moreover, its focus is on
specification and verification whereas StreamIt aims to produce an
efficient implementation.

