\section{StreamIt Overview}

StreamIt is a language for writing applications built around
continuous streams of data.  A program consists of a set of filters
that process conceptually infinite data streams; connections between
filters are implicit in the structure of the source code.

\subsection{Setting Up the Infrastructure}

StreamIt has a number of dependencies.  You need a Java runtime
environment.  The build system assumes IBM's Jikes compiler
(\url{http://www.jikes.org}).  Additionally, the C library depends on
the FFTW library (\url{http://www.fftw.org/}); you must build the
single-precision version of the library and install it somewhere in
the default compiler search path.

Unpack the StreamIt source in a directory.  This document assumes it
is unpacked in a directory named \textsf{streams} in your home
directory, but this is not a requirement.  The \textsf{STREAMIT\_HOME}
environment variable needs to be set to point to this directory.  Edit
your shell's initialization files.  If you use \textsf{bash}, edit
your \textsf{.bashrc} file and add

\begin{verbatim}
export STREAMIT_HOME=$HOME/streams
. $STREAMIT_HOME/include/dot-bashrc
\end{verbatim}

If you use \textsf{tcsh}, edit your \textsf{.cshrc} file and add

\begin{verbatim}
setenv STREAMIT_HOME=${HOME}/streams
source ${STREAMIT_HOME}/include/dot-cshrc
\end{verbatim}

Then log out and log in again, or reload your dotfiles.

To build the compiler, change directories to
\textsf{\$STREAMIT\_HOME\slash{}com\-pi\-ler} and run \textsf{make}.
There is also a C library used for the uniprocessor backend; change to
\textsf{\$STREAMIT\_HOME\slash{}library\slash{}c} and run
\textsf{make}.  You also may need to explicitly compile the files in
\textsf{\$STREAMIT\_HOME\slash{}library\slash{}java\slash{}streamit}.

\subsection{Compiling StreamIt Programs}

The \textsf{knit} script will compile a StreamIt program end-to-end.
By default, it takes the name of a file containing code as a
command-line parameter, and produces an executable called
\textsf{a.out}.  Running the executable will run the program forever;
a command-line parameter of \textsf{-i10} will run the program for ten
steady-state executions.

A StreamIt program can also be compiled and run against a Java runtime
library.  This involves using a syntax converter, and then compiling
and running the program:

\begin{verbatim}
java streamit.frontend.ToJava --library Hello.str \
  --output Hello.java
javac Hello.java
java Hello -i10
\end{verbatim}