\documentclass[11pt]{article}

\usepackage{booktabs}
%\usepackage[margin=0.5in]{geometry}
\usepackage{palatino}
\usepackage{tabularx}

% Make picture sizes be sane:
\setlength{\unitlength}{\baselineskip}

\title{A Stream Kernel Abstraction for the PCA Morphware Stream
  Virtual Machine Representation}
\author{David Maze \and William Thies \and Michael Gordon \and Michael
  Karczmarek \and Saman Amarasinghe}
\date{October 15, 2002}

\begin{document}

\maketitle
\tableofcontents

\section{Introduction}

This document describes the stream kernel language for the stream
virtual machine proposal in the Morphware forum.  This abstraction is
largely based on the MIT StreamIt language, in which incoming data is
processed incrementally as a ``stream'', taking a possibly infinite
series of inputs and producing a similar stream of outputs.

A stream kernel is a self-contained section of a larger system that
does streaming computation.  Within a kernel, the programmer
constructs a stream graph, containing blocks with a single input and a
single output, and describes the function of atomic blocks and the
structure of composite blocks.  The compiler generates code for each
block, and applies optimizations to the stream graph to produce
efficient code for the target architecture.

The kernel language is vaguely reminiscent of other imperative
languages such as C or Java.  In particular, the bodies of filter work
functions and stream initialization code is generally legal Java
code.  This provides a straightforward mapping from the kernel
language to the SAAL ISA.

\section{Types}

This section describes the various data and object types used in
stream kernels.  Data types are passed along tapes between stream
objects, and can be used as local variables.  Stream objects are only
created in initialization code and form a static stream graph.

\subsection{Data Types}

Data types are always created atomically.  They are of fixed size, and
are generally statically allocated.

\subsubsection{Primitive Types}

The following primitive types exist:

\begin{description}
\item[\texttt{bit}]  A one-bit unsigned integer type, containing
  the value \texttt{0} or \texttt{1}.

\item[\texttt{int}]  A signed integer type, of unspecified length.
  This length will typically be the native word length on the target
  machine.

\item[\texttt{float}]  A floating-point type, of the best precision
  that will give good performance.  On architectures with hardware
  support for double-precision floating-point, this will generally be
  used, but there is no guarantee of a particular bit width.

\item[\texttt{complex}]  A floating-point complex type, of the same
  precision as \texttt{float}.  This has real and imaginary parts,
  which can be directly accessed with \texttt{c.real} and
  \texttt{c.imag}.
\end{description}

\label{sec:operators}
A number of operations are supported on these primitive types; these
are listed in Table \ref{tab:primitive-operators}.  In a binary or
ternary expression, if two expressions are of different primitive
types, they are promoted to the lowest type on the list above.  Real
expressions converted to \texttt{complex} have an imaginary part of
zero.  \texttt{|}, \texttt{\&}, and \texttt{\^} are bitwise or,
and, and exclusive-or operators, respectively.  The boolean logic
operators \texttt{\&\&} and \texttt{||} behave as in C: they are
short-circuiting, integer values of exactly zero are false, and other
values are true.  \texttt{==} and other comparison operators return
an integer 1 if the comparison is true and an integer 0 if the
comparison is false.

\begin{table}
\begin{center}
\begin{tabular}{ccl}
\toprule
\textbf{Operators} & \textbf{Types} \\
\midrule
\texttt{?:} & Any & First part must be \texttt{int} or \texttt{bit} \\
\texttt{||} & \texttt{int}, \texttt{bit} \\
\texttt{\&\&} & \texttt{int}, \texttt{bit} \\
\texttt{|}, \texttt{\&}, \texttt{\^} &
  \texttt{int}, \texttt{bit} \\
\texttt{==}, \texttt{!=} & Any \\
\texttt{<}, \texttt{<=}, \texttt{>}, \texttt{>=} &
  \texttt{int}, \texttt{float} \\
\texttt{+}, \texttt{-} &
  \texttt{int}, \texttt{float}, \texttt{complex} \\
\texttt{*}, \texttt{/}, \texttt{\%} &
  \texttt{int}, \texttt{float}, \texttt{complex} &
  \texttt{\%} must be \texttt{int} \\
\texttt{++}, \texttt{--} & \texttt{int} \\
\bottomrule
\end{tabular}
\end{center}
\caption{Operators on primitive types}
\label{tab:primitive-operators}
\end{table}

\subsubsection{Structures}
\label{sec:data-structures}

Named structures of heterogeneous data types are supported.  A
structure must be of a fixed size.  It contains a set of fields, each
of which has a name and a data type.  Any primitive or array type, and
any structure type which has been previously declared in the program,
may be used; recursive structures are not allowed.  The following two
structure definitions are both legal:

\begin{verbatim}
struct A {
  int a;
  int[4] b;
}

struct B {
  A a;
  A[4] b;
}
\end{verbatim}

A structure definition contains the keyword \texttt{struct}, the name
of the structure, an open brace, a listing of field (variable)
declarations, and a close brace.  It is not followed by a semicolon.

The only supported operations on structures are field references and
copying.  Field references are of the form \texttt{a.b}; \texttt{a} must
be of some structure type \texttt{A}, and \texttt{b} must be a field named
in the structure declaration of \texttt{A}.  The type of this expression
is the declared type of the field.  Note that a similar syntax is used
for referencing the real and imaginary parts of complex numbers; there
is an implicit structure declaration

\begin{verbatim}
struct complex {
  float real;
  float imag;
}
\end{verbatim}

\noindent
along with some syntactic sugar to make complex arithmetic work.

\subsubsection{Arrays}

Arrays of any data type described here are supported.  Arrays must be
of a fixed length.  Array types are written with the base type name,
followed by a single dimension in square brackets.  Multi-dimensional
arrays or matrices are supported as arrays of arrays.

\subsubsection{Variable Declarations}

Variable declarations appear in structure declarations, as well as in
code blocks and \texttt{filter} declarations.  A variable declaration
always has the form \texttt{type name;}, where \texttt{type} is a data
type as described here.  Of note, the syntax for array declarations is
different from C and Java.  A correct declaration is
\texttt{int[4] rgbi;}, with the entire type stated before the variable name.


\subsection{Stream Types}

Computation is performed within stream objects.  Every stream object
has an input type and an output data type; the object is connected to
``tapes'' of a hypothetically infinite number of homogeneous data
objects.  \emph{Filters} are atomic, and have initialization code and
steady-state work code; \emph{pipelines}, \emph{split-joins}, and
\emph{feedback loops} are all composite structures that include some
number of stream objects as children.

All stream objects have initialization code that runs when the object
is first created.  For filters, this code may be omitted, and
generally just sets up filter-local state.  For composite objects, the
initialization code is responsible for creating the child objects.
Filters also have one or more \emph{work functions}, which are called
each time the filter executes.

\paragraph{I/O types.}  Each stream declared at the top level must
also explicitly declare its input and output types.  This is a
declaration like \texttt{int->int} or \texttt{float->complex}; the first
type is the stream's input type, the second its output type.  Either
type may be \texttt{void}; in this case, the stream has no input or
output, as appropriate.  If it is a filter, it must not declare a
peek/pop or push rate.  The top-level stream in a program has type
\texttt{void->void}; no other streams of this type are allowed.

\subsubsection{Stream Parameters}

A stream may have a set of parameters provided to it.  These
parameters are available to all functions within the stream type, and
may not be changed.  Values for these parameters must be passed in
when the stream is created.  Stream parameters are declared with types
and names in a parenthesized list after the name of the stream; when a
stream it instantiated, values for the parameters must follow the name
of the stream.  For example:

\begin{verbatim}
float->float filter MatrixMatrixMultiply(int A,
                 int B, int C) { ... }
add MatrixMatrixMultiply(3, 5, 4);
\end{verbatim}

\paragraph{Compile-time constants.}  Several things must
be \emph{compile-time constant}; for example, the I/O rates of a
filter must be determined at compile-time, and the compiler must be
able to statically create the expanded stream graph.  An expression is
compile-time constant if it is a literal, a stream parameter, or an
expression whose components are entirely compile-time constant.

\subsubsection{Parameterized Stream Types}

Certain built-in stream types may be \emph{parameterized} on some data
type or types, much like ``template'' types in C++.  These are
detailed in section \ref{sec:built-in-streams}; they cannot be created
by the user.  To instantiate a built-in parameterized type, use the
name of the stream type, followed by the name of the data type in
angle brackets, such as \texttt{Identity<int>}.

\subsubsection{Filters}

All computation in a stream kernel takes place within filters.
Filters must explicitly declare their initialization code if present,
but also have the option of omitting it.  A filter may have a single
anonymous work function, or multiple named work functions.  Each work
function must declare its \emph{I/O rates}, the number of items one
call to the function removes, examines, or places on its tape.  A
basic filter declaration looks like this:

\begin{verbatim}
int->int filter IntAvgFilter {
  init {
    // empty
  }
  work pop 1 peek 2 push 1 {
    push(peek(0) + peek(1));
    pop();
  }
}
\end{verbatim}

This filter is named \texttt{IntAvgFilter}; it reads integers off of its
input tape, and writes integers to its output tape.  Each execution of
the work function removes exactly one item from the input tape and
examines at most two items; it causes exactly one item to be pushed on
to the output tape.

\paragraph{Init functions.}  Init functions are declared using the
keyword \texttt{init}.  They take no parameters.  They may initialize
filter fields or do other work, but have no access to the filter's
tapes.

\paragraph{Work functions.}  Work functions are declared using the
keyword \texttt{work}, which may be followed by a name.  At most
one work function may be declared anonymously.  The function name must
be followed by rate declarations, of the form \texttt{push expr},
\texttt{peek expr}, and \texttt{pop expr}, in any order.  If the
peek rate is omitted, it is assumed to be equal to the pop rate; if
the pop or push rate is omitted, it is assumed to be zero, and the
corresponding input or output tape type must be \texttt{void}.  The
function must push and pop exactly the number of items specified, and
may peek up to the number of items specified.  All of the rate
expressions must be compile-time constant.

\paragraph{Filter state.}  The top level of the filter may include
variable declarations.  These variables are visible to every function
in the filter, and may be used to carry state between iterations of
the work function.


\subsubsection{Pipelines}

The simplest composite filter is a pipeline.  A pipeline contains a
number of child streams; the output of the first stream is connected
to the input of the second, whose output is connected to the input of
the third, and so on.  A pipeline declaration looks like

\begin{verbatim}
float->complex pipeline FloatToComplexAdd {
  add ReImToComplex();
  add ComplexPairAdd();
}
\end{verbatim}

The body of the pipeline declaration is the initialization code; no
internal function declarations or message handlers are allowed.  In
addition to the internal typing constraints mentioned previously, the
input type of the pipeline must match the input type of the first
filter, and the output type of the pipeline must match the output type
of the last filter.

\begin{figure}[htbp]
  \begin{center}
    \begin{picture}(18,9)
      \put(3,9){\vector(0,-1){1}}
      \put(0,1){\framebox(6,7)[lt]{\texttt{Pipeline}}}
      \put(7,8){\makebox(9,1)[l]{\texttt{float->float}}}
      \put(7,7){\makebox(9,1)[l]{\texttt{pipeline Pipeline \{}}}
      \put(3,8){\vector(0,-1){1}}
      \put(1,6){\framebox(4,1){\texttt{Child1}}}
      \put(7,6){\makebox(9,1)[l]{\texttt{\ add Child1();}}}
      \put(3,6){\vector(0,-1){1}}
      \put(1,4){\framebox(4,1){\texttt{Child2}}}
      \put(7,4){\makebox(9,1)[l]{\texttt{\ add Child2();}}}
      \put(3,4){\vector(0,-1){1}}
      \put(1,2){\makebox(4,1){$\cdots$}}
      \put(7,2){\makebox(9,1)[l]{\texttt{\ ...}}}
      \put(3,2){\vector(0,-1){1}}
      \put(7,1){\makebox(9,1)[l]{\texttt{\}}}}
      \put(3,1){\vector(0,-1){1}}
    \end{picture}
    \caption{Pipelines}
    \label{fig:pipeline}
  \end{center}
\end{figure}


\subsubsection{Split-Joins}

A split-join allows computation to be run in parallel, possibly using
different parts of the input stream.  It is so named because incoming
data passes through a \emph{splitter}, is redistributed to the child
streams for processing, and then is fed through a \emph{joiner} to be
recombined into a single output stream; see section
\ref{sec:expr-splitters-joiners}.

\begin{verbatim}
int->int splitjoin AddAndSub {
  split duplicate;
  add IntAdder();
  add IntSubtractor();
  join roundrobin;
}
\end{verbatim}

This split-join has two children.  Incoming data is duplicated on to
both streams.  The first branch consumes two elements and adds them;
the second consumes two elements and subtracts them.  The resulting
values are then pushed on to the output stream in a round-robin
fashion, with values from alternating streams.

\begin{figure}[htbp]
  \begin{center}
    \begin{picture}(29,5)
      \put(6,5){\vector(0,-1){1}}
      \put(0,1){\framebox(12,3)[tl]{\texttt{SplitJoin}}}
      \put(13,4){\makebox(16,1)[l]
        {\texttt{int->int splitjoin SplitJoin \{}}}
      \put(6,4){\vector(-3,-1){3}}
      \put(6,4){\vector(0,-1){1}}
      \put(6,4){\vector(3,-1){3}}
      \put(13,3){\makebox(16,1)[l]{\texttt{\ split roundrobin(2);}}}
      \put(1,2){\framebox(4,1){\texttt{Child1}}}
      \put(5,2){\makebox(2,1){$\cdots$}}
      \put(7,2){\framebox(4,1){\texttt{ChildN}}}
      \put(13,2){\makebox(16,1)[l]
        {\texttt{\ add Child1(); ... add ChildN();}}}
      \put(3,2){\vector(3,-1){3}}
      \put(6,2){\vector(0,-1){1}}
      \put(9,2){\vector(-3,-1){3}}
      \put(13,1){\makebox(16,1)[l]{\texttt{\ join roundrobin;}}}
      \put(6,1){\vector(0,-1){1}}
      \put(13,0){\makebox(16,1)[l]{\texttt{\}}}}
    \end{picture}
    \caption{Split-Joins}
    \label{fig:splitjoin}
  \end{center}
\end{figure}

\subsubsection{Feedback Loops}

A feedback loop has a body stream.  Its output passes through a
splitter; one branch of the splitter leaves the loop, and the other
goes to the loop stream.  The output of the loop stream and the loop
input then go through a joiner to the body's input.

\begin{verbatim}
float->float feedbackloop AddFeedback(float scaling) {
  join roundrobin;
  body FloatAdder();
  loop FloatScaler(scaling);
  split duplicate;
  enqueue 0.0;
}
\end{verbatim}

\paragraph{Semantics of splitters and joiners.}  Splitters and joiners
both treat the stream input as the first child, and the loop edge as
the second child.  Splitters and joiners that do not allow exactly two
children are disallowed.

\paragraph{\texttt{enqueue} statement.}  The \texttt{enqueue}
statement takes a value and places it in FIFO order on the output tape
from the loop stream.  You will generally need to enqueue enough items
that the input joiner will be able to fire once; enqueuing more items
causes a delay in the feedback loop.

\begin{figure}[htbp]
  \begin{center}
    \begin{picture}(21,9)
      \put(6,9){\vector(0,-1){1}}
      \put(10,8){\makebox(11,1)[l]{\texttt{float->float}}}
      \put(0,1){\framebox(9,7)[lt]{\texttt{Feedback}}}
      \put(10,7){\makebox(11,1)[l]{\texttt{feedbackloop Feedback \{}}}
      \put(6,8){\vector(0,-1){1}}
      \put(10,6){\makebox(11,1)[l]{\texttt{\ join roundrobin;}}}
      \put(6,7){\vector(0,-1){1}}
      \put(4,5){\framebox(4,1){\texttt{Body}}}
      \put(10,5){\makebox(11,1)[l]{\texttt{\ body Body();}}}
      \put(6,5){\vector(0,-1){3}}
      \put(6,2){\line(-1,0){3}}
      \put(3,2){\vector(0,1){1}}
      \put(1,3){\framebox(4,1){\texttt{Loop}}}
      \put(10,3){\makebox(11,1)[l]{\texttt{\ loop Loop();}}}
      \put(3,4){\line(0,1){3}}
      \put(3,7){\vector(1,0){3}}
      \put(6,2){\vector(0,-1){1}}
      \put(10,1){\makebox(11,1)[l]{\texttt{\ split roundrobin(4,1);}}}
      \put(6,1){\vector(0,-1){1}}
      \put(10,0){\makebox(11,1)[l]{\texttt{\}}}}
    \end{picture}
    \caption{Feedback Loop}
    \label{fig:feedback-loop}
  \end{center}
\end{figure}

\subsubsection{Anonymous Streams}
\label{sec:anonymous-streams}

It is possible to declare a stream object without a name, if it is
used in exactly one place in the program.  In general anonymous stream
declarations look exactly like the corresponding normal stream
declaration, except that the stream type and name are omitted.

\begin{verbatim}
complex->float feedbackloop FeedbackMagnitude(float scaling)
{
  join roundrobin;
  body filter {
    work pop 2 push 1 { push(abs(peek(0) * peek(1))); pop(); pop(); }
  };
  loop pipeline {
    add FloatScaler(scaling);
    add FloatToComplex();
  };
  split duplicate;
}
\end{verbatim}

The input and output types of anonymous streams are determined by the
compiler.

\section{Statements}

This section describes statements legal in stream kernel code.  For
the most part, these are identical to legal statements in C or Java.
Some statements, particularly those that set up the stream graph, are
only legal in initialization code; other statements are legal
anywhere.

\subsection{Initialization-Only Statements}

\subsubsection{\texttt{add} Statement}

The \texttt{add} statement adds a child stream to the current stream
object.  It is valid in split-joins and pipelines only.  In a
split-join, it adds a new stream to the end of the current list of
children.  In a pipeline, it adds a new stream to the end of the chain
of children.

This statement takes a \emph{stream constructor}; see section
\ref{sec:expr-stream-constructor}.

\subsubsection{\texttt{body} Statement}

The \texttt{body} statement adds a child stream as the body part of
a feedback loop.  This statement must be executed exactly once in the
initialization code of a feedback loop, and is invalid anywhere else.

This statement takes a \emph{stream constructor}; see section
\ref{sec:expr-stream-constructor}.

\subsubsection{\texttt{enqueue} Statement}

The \texttt{enqueue} statement is valid only within the
initialization code of a feedback loop.  It causes the specified item
to be placed after all other previously enqueued items on the input to
the feedback loop joiner.  This statement takes a single value
expression.

\subsubsection{\texttt{join} Statement}

The \texttt{join} statement declares the joiner type of a feedback
loop or split-join.  It only appears in initialization code, and is
invalid in initialization code for any other construct.  This
statement takes a joiner specification; the only one valid currently
is \texttt{roundrobin}, see section \ref{sec:expr-round-robin}.

\subsubsection{\texttt{loop} Statement}

The \texttt{loop} statement adds a child stream as the loop part of
a feedback loop.  This statement must be executed exactly once in the
initialization code of a feedback loop, and is invalid anywhere else.

This statement takes a \emph{stream constructor}; see section
\ref{sec:expr-stream-constructor}.

\subsubsection{\texttt{split} Statement}

The \texttt{split} statement declares the splitter type of a feedback
loop or split-join.  It only appears in initialization code, and is
invalid in initialization code for any other construct.  This
statement takes a splitter specification; only \texttt{duplicate} and
\texttt{roundrobin} are currently legal, see sections
\ref{sec:expr-duplicate} and \ref{sec:expr-round-robin}.


\subsection{Work-only Statements}

These statements may only appear within steady-state work code, and in
particular, within a filter's \texttt{work} function(s).

\subsubsection{\texttt{push} Statement}

The \texttt{push} statement takes a value expression and pushes its
value on to the output tape of the current filter.  The type of the
expression must match the output type of the filter.

\subsection{General-use Statements}

These statements may appear anywhere code is legal.

\subsubsection{Statements with Java-like Semantics}

\texttt{do}, \texttt{for}, and \texttt{while} loops have the same
syntax and semantics they do in Java.  The \texttt{continue} and
\texttt{break} statements are also recognized, though they may not
take  a label reference.

For statements whose syntax match exactly

\begin{verbatim}
for (i = S; i < N; i++)
  /* body */ ;
\end{verbatim}

\noindent
cause the induction variable \texttt{i} to be compile-time constant
within the loop if the start and end indices \texttt{S} and \texttt{N} are
also compile-time constant.

\subsubsection{Assignment Statements}

Assignment statements have the form \texttt{lhs = expr;}.  The left-hand
side may be a variable, a field reference, or an array element
reference.  Each of these is evaluated to a location, with the result
being either a primitive value, a structure, or an entire array.  If
it is a primitive value, the right-hand side must be a value
expression.  Otherwise, the right-hand side must be an array or
structure of identical type, and causes all of the elements from the
right-hand-side object to be copied to the left-hand-side object.

\subsubsection{Variable Declarations}

Variable declarations have the form \texttt{type name;}.  \texttt{type}
must be a data type; \texttt{name} may be any legal name not
corresponding to another variable previously declared in the current
block.  A declaration must appear before the variable is used.  The
variable declaration may also contain an initialization, as in
\texttt{type name = expr;}  The initialization acts exactly as an
initialization statement, above.

Variables declared without initialization are initialized to 0 as best
as possible.  \texttt{complex} variables are initialized to
\texttt{0+0i}.  Each element of an array variable is initialized to
0 as described here.  Similarly, each field of a structure variable is
initialized to 0 in the same way.


\section{Expressions}

\subsection{Value Expressions}
\label{sec:expr-value}

A \emph{value expression} carries some value of a data type.  Literal
values are value expressions; these can be signed integer or real
values, e.g. \texttt{17}, \texttt{2.45}, or \texttt{1.4e6}.  A literal value
can also be a pure imaginary number, e.g. \texttt{17i} or \texttt{3.4i}.
Unary, binary, and ternary expressions of value expressions are also
value expressions, as described in section \ref{sec:operators}.  Thus,
while \texttt{3+4i} is not a literal, it is a valid value expression of
complex type.

Field references to structure objects are also value expressions.
These have the form \texttt{a.b}, where \texttt{a} is a variable of
structured type \texttt{A} and \texttt{b} is a field in \texttt{A}.
Similarly, array references \texttt{a[b]} are value expressions;
\texttt{a} must be an array variable, and \texttt{b} must be an
integer-valued value expression.

\subsubsection{\texttt{peek} and \texttt{pop}}
\label{sec:expr-peek-pop}

The \texttt{peek} and \texttt{pop} expressions look at items on
the incoming tape in a filter's work function; they are illegal in any
other context.  \texttt{peek(n)} examines the n-th item on the tape
without removing it; \texttt{peek(0)} returns the next incoming
item, \texttt{peek(1)} the one afterwards, and so on.
\texttt{pop()} returns the same item \texttt{peek(0)} returns,
but also removes it from the tape.

Note that \texttt{pop()} removes an item from the tape, so
\texttt{peek()}s afterwards use different indices.  A work function
must pop a fixed number of items, and declare the number of items in
its rate declaration.  It may not peek at more items than it declares
in its rate declaration, though it may peek at fewer.  The peek
declaration declares the furthest position on the tape examined from
the point of view of the beginning of the function.

\begin{verbatim}
int->int filter PeekRates {
  work pop 1 peek 3 push 1 {
    int a, b, c, d, e;
    a = peek(0);
    b = peek(1);
    c = pop();   // Same value as a
    d = peek(0); // Same value as b
    e = peek(1);
    push(a + b + c + d + e);
  }
}
\end{verbatim}

This function pops one item and pushes 1.  It declares a peek rate of
3 because \texttt{e} peeks at the third item that was on the tape
at the start of the function -- one is popped, and \texttt{e}
contains the second item beyond that.

\subsubsection{Function Calls}
\label{sec:expr-funcall}

Function calls are value expressions; certain cases of built-in
functions may return values that are compile-time constant if their
parameters are compile-time constant.  The function call syntax
\texttt{fn(p1, p2)} is identical to that of Java.  All parameters are
always passed by value; this differs from Java's semantics, in which
objects and arrays are passed by reference.

Function calls may be made to declared helper functions, or to
built-in math functions.  The math functions are listed in table
\ref{tab:math-functions}.

\begin{table}[htbp]
  \begin{center}
    \begin{tabular}{cl}
      \toprule
      \texttt{abs(v)} & Real absolute value or complex magnitude \\
      \texttt{arg(v)} & Polar angle of complex value \\
      \texttt{exp(v)} & Real or complex exponent \\
      \texttt{log(v)} & Real or complex natural logarithm \\
      \texttt{sin(v)} & Real or complex sine \\
      \texttt{cos(v)} & Real or complex cosine \\
      \texttt{sqrt(v)} & Real or complex square root \\
      \texttt{csqrt(v)} & Complex square root \\
      \bottomrule
    \end{tabular}
    \caption{Built-In Math Functions}
    \label{tab:math-functions}
  \end{center}
\end{table}

\subsection{Stream Constructors}
\label{sec:expr-stream-constructor}

A \emph{stream constructor} indicates that a new stream object should
be created.  Stream constructors appear in \texttt{add},
\texttt{body}, and \texttt{loop} statements in composite stream
initialization code.

Stream constructors may reference an already-declared stream type.  In
this case, the stream constructor form is \texttt{Name(param, ...)}.
There must be exactly as many parameters in the constructor as
parameters for the stream type.  The parameters must all be
compile-time constant.

An anonymous stream declaration is also a valid stream constructor;
see section \ref{sec:anonymous-streams}.

\subsection{Splitters and Joiners}
\label{sec:expr-splitters-joiners}

Splitters have a single input of a given type and multiple outputs of
the same type; joiners have multiple inputs and a single output.  They
appear in feedback loops and split-joins.

\subsubsection{\texttt{duplicate} Splitters}
\label{sec:expr-duplicate}

\texttt{duplicate} splitters take each incoming item and push the
same item to each of the outgoing tapes, duplicating data.

\subsubsection{\texttt{roundrobin} Splitters and Joiners}
\label{sec:expr-round-robin}

\texttt{roundrobin} splitters take each item and send it to exactly
one of the child streams, in order.  Either \texttt{roundrobin} or
\texttt{roundrobin()} causes one item to be sent to each output, in
order; \texttt{roundrobin(2)} causes two items to be sent to the first
stream, two to the second, and so on.  \texttt{roundrobin(2, 4, 2)}
requires there to be exactly three children, and sends two to the
first child, four to the second, and two to the third.  A round-robin
weight may be 0; in this case, the child input must be of type
\texttt{void}.

\texttt{roundrobin} joiners are identical to \texttt{roundrobin}
splitters, except that they read from the input tapes in the specified
pattern and write data to the output tape.


\section{Built-in Types}

This section describes built in data and stream types.  These types
are implicitly declared; code can just use these types without
explicitly importing or otherwise declaring them.

\subsection{Built-in Data Types}

The \texttt{complex} type is actually implemented as a structure,
though the compiler implements syntactic sugar for complex variables.
The structure definition is shown above in section
\ref{sec:data-structures}.
It is legal to explicitly reference the \texttt{real} and
\texttt{imag} fields of complex numbers as though they were the
structure type instead.

\subsection{Built-in Streams}
\label{sec:built-in-streams}

Several stream types are in common use, and are either simple enough
or require non-standard language functionality that the compiler
should support them directly.  These include parameterized
\texttt{Identity}, \texttt{StreamReader}, and \texttt{StreamWriter}
types, as shown in figure \ref{fig:built-in-streams}.

\begin{figure}[htbp]
    \begin{verbatim}
template<T> T->T filter Identity {
  work pop 1 push 1 { push(pop()); }
}
template<T> void->T filter StreamReader(int n) {
  work push 1 { push(_IN(n)); }
}
template<T> T->void filter StreamWriter(int n) {
  work pop 1 { _OUT_INORDER(n, pop()); }
}
\end{verbatim}
    \caption{Built-in Stream Types}
    \label{fig:built-in-streams}
\end{figure}

\section{Mapping to the Morphware Stream Virtual Machine}

This section describes the general flow of the compiler and how a
stream kernel interacts with the Morphware stream virtual machine.
The compiler will ultimately generate C code using the \texttt{\_IN},
\texttt{\_OUT\_INORDER}, and \texttt{\_PEEK} macros to perform stream
I/O; a program can be fused into a single kernel, or transformed into
multiple kernels that communicate using these macros.

\paragraph{\texttt{StreamInput} and \texttt{StreamOutput} filters.}
The stream kernel language provides two built-in filters,
\texttt{StreamInput} and \texttt{StreamOutput}, that communicate with
the controlling threaded processor.  If the VM API macros were exposed
in the stream language, these could be implemented as shown in figure
\ref{fig:built-in-streams}.

These types exist such that it is possible to programmatically
determine the top-level stream in a kernel, and so that inputs and
outputs can be connected at arbitrary locations in the stream graph.
The top-level stream always has the type \texttt{void->void}; it can
use a \texttt{StreamInput} as a source and a \texttt{StreamOutput} as
a sink in the normal case.  Figure \ref{fig:interleave-input} shows
how \texttt{StreamInput} can also be used to interleave an input with
other data.

\begin{figure}[htbp]
  \begin{verbatim}
splitjoin InterleaveInputInt(int n)
{
  split roundrobin(1, 0);
  add Identity<int>;
  add StreamInput<int>(n);
  join roundrobin(1, 1);
}
\end{verbatim}
  \caption{Interleaving input with data}
  \label{fig:interleave-input}
\end{figure}

\paragraph{Mapping to multiple stream kernels.}  The stream compiler
can apply fusion and fission transformations to turn a stream program
into multiple stream kernels.  In this case, the compiler will
automatically generate appropriate calls to \texttt{\_IN} and
\texttt{\_OUT\_INORDER} to communicate between streams.  The compiler
will also generate the stub code required on the threading processor
to set up, execute, and reap the streams.

\end{document}