
\title{Stream Programming for Image and Video Compression}

\author{Matthew Henry Drake}
\department{Department of Electrical Engineering and Computer~Science}
% If the thesis is for two degrees simultaneously, list them both
% separated by \and like this:
% \degree{Doctor of Philosophy \and Master of Science}
\degree{Bachelor of Science in Computer Science and Engineering
        \and Master of Engineering in Electrical Engineering and Computer Science}
\degreemonth{May}
\degreeyear{2006}
\thesisdate{May 26, 2006}

%% By default, the thesis will be copyrighted to MIT.  If you need to copyright
%% the thesis to yourself, just specify the `vi' documentclass option.  If for
%% some reason you want to exactly specify the copyright notice text, you can
%% use the \copyrightnoticetext command.  
%\copyrightnoticetext{\copyright IBM, 1990.  Do not open till Xmas.}

% If there is more than one supervisor, use the \supervisor command
% once for each.
\supervisor{Saman Amarasinghe}{Associate Professor}

% This is the department committee chairman, not the thesis committee
% chairman.  You should replace this with your Department's Committee
% Chairman.
\chairman{Arthur C. Smith}{Chairman, Department Committee on Graduate Theses}

% Make the titlepage based on the above information.  If you need
% something special and can't use the standard form, you can specify
% the exact text of the titlepage yourself.  Put it in a titlepage
% environment and leave blank lines where you want vertical space.
% The spaces will be adjusted to fill the entire page.  The dotted
% lines for the signatures are made with the \signature command.
\maketitle

% The abstractpage environment sets up everything on the page except
% the text itself.  The title and other header material are put at the
% top of the page, and the supervisors are listed at the bottom.  A
% new page is begun both before and after.  Of course, an abstract may
% be more than one page itself.  If you need more control over the
% format of the page, you can use the abstract environment, which puts
% the word "Abstract" at the beginning and single spaces its text.

%% You can either \input (*not* \include) your abstract file, or you can put
%% the text of the abstract directly between the \begin{abstractpage} and
%% \end{abstractpage} commands.

% First copy: start a new page, and save the page number.
\cleardoublepage
% Uncomment the next line if you do NOT want a page number on your
% abstract and acknowledgments pages.
% \pagestyle{empty}
\setcounter{savepage}{\thepage}
\begin{abstractpage}
\input{abstract}
\end{abstractpage}

% Additional copy: start a new page, and reset the page number.  This way,
% the second copy of the abstract is not counted as separate pages.
% Uncomment the next 6 lines if you need two copies of the abstract
% page.
% \setcounter{page}{\thesavepage}
% \begin{abstractpage}
% \input{abstract}
% \end{abstractpage}

\cleardoublepage

\section*{Acknowledgments}

\begin{quote}
\noindent\textit{Though we are not now that strength which in old days\\
Moved earth and heaven; that which we are, we are;\\
One equal temper of heroic hearts,\\
Made weak by time and fate, but strong in will\\
To strive, to seek, to find, and not to yield.
}

Tennyson, \textit{Ulysses} (1842)
\end{quote}

My thanks go first to Rodric Rabbah and William Thies for 
their direction, advice, and innumerable hours of help; I could
not have asked for better role models and guides to the world
of computer science research. 
I thank my advisor Saman Amarasinghe for supporting and guiding
this thesis and demanding the highest of standards.

I am grateful to everyone in the StreamIt group, including William Thies, 
Rodric Rabbah, Michael Gordon, Allyn Dimock, and Janis Sermulins,
for their extensive work on the StreamIt language and compiler. Without
their help this thesis would not have been possible.

I want to thank my parents for twenty two years spent instilling in me the 
importance of education and hard work - particularly, my father's
early encouragement of my programming interests and my mother's 
demand for impeccable writing. 
I owe a huge debt of gratitude to my grandfather, Benjamin Turner, 
and Meng Mao, for their encouragement and interest in my work.

This thesis extends from a paper\cite{drake:ipdps:2006} by
the author, Rodric Rabbah, and Saman Amarasinghe that appeared in
the proceedings of the 
IEEE 2006 International Parallel and Distributed Processing
Symposium. 
The StreamIt project is supported by DARPA grants 
PCA-F29601-03-2-0065 and HPCA/PERCS-W0133890, and
NSF awards CNS-0305453 and EIA-0071841.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% -*-latex-*-
