\section{Introduction}

Recent years have seen the proliferation of applications that are
based on some notion of a ``stream''.  There is evidence that
streaming media applications are already consuming most of the cycles
on consumer machines \cite{Rix98}, and their use is continuing to
grow.  The stream abstraction is central to embedded applications for
hand-held computers, cell phones, and DSP's, as well as for
high-performance applications such as intelligent software routers,
cell phone base stations, and HDTV editing consoles.

Despite the prevalence of these applications, there is surprisingly
little language and compiler support for practical, large-scale stream
programming.  For example, a number of grid-based architectures have
been emerging that are particularly well-suited for stream programming
\cite{raw, smartmemories, trips}, but there is no common machine
language that a programmer can use to exploit their common properties
while hiding their differences.  Thus, most programmers turn to
general-purpose languages such as C or C++ to implement stream
programs, resorting to low-level assembly codes for
performance-critical loops.  This practice is labor-intensive,
error-prone, and very costly, since the performance-critical sections
must be re-implemented for each target architecture.  Moreover,
general purpose languages do not provide a natural and intuitive
representation of streams, thereby having a negative effect on
readability, robustness, and programmer productivity.

StreamIt is a language and compiler specifically designed for modern
stream programming.  Its goal is to raise the abstraction level in the
streaming domain, providing a natural, high-level syntax that conceals
architectural details without sacrificing performance.  To accomplish
this goal, the compiler needs to be ``stream-aware''--that is, it
needs to be able to recognize, analyze, and manipulate data streams as
would an expert assembly programmer in lowering an application to a
given target.  Towards this end, this paper makes the following
contributions:
\begin{itemize}

\item ``Structured'' streams as a language construct for enabling
  novel compiler analyses of stream programs.
  
\item The identification of a fundamental property of a stream
  graph--the stream dependence function--that establishes a notion of
  relative time and dependence information.

\item A semantic model of structured stream programs that allows one
  to formulate and validate stream transformations.

\item A parallel fusion transformation that collapses several filters
  into one.

\item A suite of optimizations that are specific to the streaming
  domain.

\item A novel phased scheduling algorithm that finds a minimal latency
schedule over a structured stream graph.

\item A prototype implementation of the StreamIt optimizing compiler
  that is showing promising results.

\end{itemize}

