Image and video codecs are prevalent in multimedia devices, ranging
from embedded systems, to desktop computers, and high-end servers such
as HDTV editing consoles. It is not uncommon however that developers
create (from scratch) and customize their codec implementations for
each of the architecture targets they intend their coders and decoders
to run on. This practice is time consuming and error prone, leading to
code that is not malleable or portable. This paper describes an
implementation of the MPEG-2 decoder using the StreamIt programming
language. StreamIt is an architecture-independent stream language that
aims to improve programmer productivity, while concomitantly exposing
the inherent parallelism and communication topology of the
application. The paper shows that MPEG is a good match for the
streaming programming model, and illustrate the malleability of the
implementation using a simple modification to the decoder to support
alternate color compression formats. StreamIt allows for modular
application development, which increases code reuse, and reduces the
complexity of the debugging process since stream components can be
verified independently. This in turn leads to greater programmer
productivity.
%% We implemented a fully functional MPEG-2 decoder in StreamIt. The
%% decoder was developed in eight weeks by a single student programmer
%% who did not have any prior experience with MPEG or other video codecs.
%% Many of the MPEG-2 components were subsequently reused to
%% assemble a JPEG codec.
