\Section{Concluding Remarks}

In this paper we described our MPEG-2 codec implementation as it was
developed using the StreamIt programming language. Our MPEG-2 decoder
was developed in eight week by a single student programmer with no
prior MPEG knowledge. We showed how the implementation is malleable by
describing how the decoder is modified to support two different chroma
sampling rates. In addition, we showed that the StreamIt language is a
good match for representing the MPEG stream flow in that there is
great correlation between a block level diagram describing the flow of
data between computation elements and the application
syntax. Furthermore, we illustrated that teleport messaging, which
allows for out-of-band communication of stream parameters, allows the
decoder to decouple the regular flow of data from the irregular
communication of decoding parameters (e.g., quantization
coefficients). This in turns leads to a cleaner implementation that is
easier to maintain and evolve with changing software specification.
In addition, we have prototypes an MPEG-2 encoder, and our current
focus is geared toward augmenting the implementation using various
motion estimation techniques.

As computer architectures change from the traditional monolithic
processors, to scalable wire-exposed and multi-core processors, there
will be a greater need for portable codec implementations that boost
programmer productivity---while also exposing parallelism and
communication to enable efficient and high performance executions.
StreamIt makes a great stride toward this end by providing a language
that features hierarchical, modular, malleable, and portable streams.