\documentclass[times,10pt,twocolumn]{article}
\usepackage{ipdps}
\usepackage{vmargin}
\setpapersize{USletter}
\setmargnohf{0.8125in}{1in}{6.875in}{8.875in}
\pagestyle{empty}
\newenvironment{twoaffiliations}{\begin{tabular}{cc}}{\end{tabular}}
\newenvironment{oddaffiliation}{\begin{center}}{\\~\\\end{center}}
%\documentclass[times,11pt,onecolumn]{article} 
%\usepackage{latex8}
%\usepackage{times}
\usepackage{epsfig}
%\pagestyle{empty}
%\newcommand{\mt}[1]{\mbox{#1}}

\title{MPEG-2 Decoding in a Stream Programming Language}

\author{
  Matthew Drake, Hank Hoffmann, Rodric Rabbah, and Saman Amarasinghe\\
  \begin{twoaffiliations}
    Massachusetts Institute of Technology\\
    Computer Science and Artificial Intelligence Laboratory\\
    \{madrake, hank, rabbah, saman\}@mit.edu
  \end{twoaffiliations}
}

\begin{document}
  
  \maketitle
  \thispagestyle{empty}
  
  \begin{abstract}
    \input{abstract}
  \end{abstract}
  
  \input{intro}
  \input{mpeg}
  \input{language}
  \input{decoder}
  \input{chroma}
  %\input{compiler}
  %\input{encoder}
  \input{related}
  %\input{future}
  \input{conclusion}
  
  \vspace{-2pt}
  \section*{Acknowledgements}
  \vspace{-7pt}
  We are very grateful to the entire StreamIt team for their hard work
  and insightful comments. Allyn Dimock, Michael Gordon, Janis
  Sermulins, and especially William Thies, contributed immensely to
  the StreamIt infrastructure to enable this paper. We also thank the
  anonymous reviewers for their helpful suggestions. The StreamIt
  project is supported by DARPA grants PCA-F29601-03-2-0065 and
  HPCA/PERCS-W0133890, and NSF awards CNS-0305453 and EIA-0071841.

%  \bibliographystyle{ipdps}
%  \bibliography{main}
  \vspace{-2pt}
\begin{thebibliography}{10}\setlength{\itemsep}{-1ex}\small
  \vspace{-7pt}
\bibitem{agrawal05cases}
S.~Agrawal, W.~Thies, and S.~Amarasinghe.
\newblock Optimizing stream programs using linear state space analysis.
\newblock In {\em {CASES}}, 2005.

\bibitem{ahmad01multiproc}
I.~Ahmad, S.~M. Akramullah, M.~L. Liou, and M.~Kafeel.
\newblock {A Scalable Off-line MPEG-2 Video Encoding Scheme using a
  Multiprocessor System}.
\newblock {\em {Parallel Computing}}, 27, 2001.

\bibitem{ahmad01compression}
I.~Ahmad, Y.~He, and M.~L. Liou.
\newblock {Video compression with parallel processing}.
\newblock {\em {Parallel Computing}}, 28, 2002.

\bibitem{ph}
S.~Aidtya, Arvind, L.~Augustsson, J.~Maessen, and R.~S. Nikhil.
\newblock {Semantics of pH: A parallel dialect of Haskell}.
\newblock In {\em Haskell Workshop}, 1995.

\bibitem{Lucid77}
E.~Ashcroft and W.~Wadge.
\newblock Lucid, a non procedural language with iteration.
\newblock {\em C. ACM}, 20(7), 1977.

\bibitem{assayad05mpeg4b}
I.~Assayad, P.~Gerner, S.~Yovine, and V.~Bertin.
\newblock {Modelling, Analysis and Parallel Implementation of an On-line Video
  Encoder}.
\newblock In {\em {1st Int. Conf. on Distributed Frameworks for Multimedia
  Applications}}, 2005.

\bibitem{Esterel}
G.~Berry and G.~Gonthier.
\newblock {The Esterel Synchronous Programming Language: Design, Semantics,
  Implementation}.
\newblock {\em Sci. of Comp. Programming}, 19(2), 1992.

\bibitem{brook04}
I.~Buck, T.~Foley, D.~Horn, J.~Sugerman, K.~Fatahalian, M.~Houston, and
  P.~Hanrahan.
\newblock {Brook for GPUs: Stream Computing on Graphics Hardware}.
\newblock In {\em SIGGRAPH}, 2004.

\bibitem{Lucid-Synchrone}
P.~Caspi and M.~Pouzet.
\newblock {Lucid Synchrone distribution}.
\newblock {\tt http://www-spi.lip6.fr/lucid-synchrone/}.

\bibitem{spidle03}
C.~Consel, H.~Hamdi, L.~Réveillère, L.~Singaravelu, H.~Yu, and C.~Pu.
\newblock {Spidle: A DSL Approach to Specifying Streaming Application}.
\newblock In {\em {2nd Int. Conf. on Generative Prog. and Component
  Engineering}}, 2003.

\bibitem{Occam}
I.~Corporation.
\newblock {\em Occam 2 Reference Manual}.
\newblock Prentice Hall, 1988.

\bibitem{kock02jpeg}
E.~de~Kock.
\newblock {Multiprocessor Mapping of Process Networks: A JPEG Decoding Case
  Study}.
\newblock In {\em {15th Int. Symp. on System Synthesis}}, 2002.

\bibitem{kock00yapi}
E.~de~Kock, G.~Essink, W.~Smits, P.~van~der Wolf, J.~Brunel, W.~Kruijtzer,
  P.~Lieverse, and K.~Vissers.
\newblock {YAPI: Application Modeling for Signal Processing Systems}.
\newblock In {\em {Conf. on Design Automation}}, 2000.

\bibitem{dwivedi01exploring}
B.~K. Dwivedi, J.~Hoogerbrugge, P.~Stravers, and M.~Balakrishnan.
\newblock {Exploring design space of parallel realizations: MPEG-2 decoder case
  study}.
\newblock In {\em {9th Int. Symp. on Hardware/Software Codesign}}, 2001.

\bibitem{sisal}
J.~Gaudiot, W.~Bohm, T.~DeBoni, J.~Feo, and P.~Mille.
\newblock {The Sisal Model of Functional Programming and its Implementation}.
\newblock In {\em {2nd Aizu Int. Symposium on Parallel Algorithms/Architecture
  Synthesis}}, {1997}.

\bibitem{Signal}
T.~Gautier, P.~L. Guernic, and L.~Besnard.
\newblock Signal: A declarative language for synchronous programming of
  real-time systems.
\newblock {\em Springer Verlag LNCS}, 274, 1987.

\bibitem{gordon02asplos}
M.~Gordon, W.~Thies, M.~Karczmarek, J.~Lin, A.~S. Meli, C.~Leger, A.~A. Lamb,
  J.~Wong, H.~Hoffman, D.~Z. Maze, and S.~Amarasinghe.
\newblock {A Stream Compiler for Communication-Exposed Architectures}.
\newblock In {\em {ASPLOS}}, 2002.

\bibitem{Lustre}
N.~Halbwachs, P.~Caspi, P.~Raymond, and D.~Pilaud.
\newblock The synchronous data flow language {LUSTRE}.
\newblock {\em Proc. of the IEEE}, 79(1), 1991.

\bibitem{MPEG2}
{ISO/IEC 11172: Information technology --- Coding of moving pictures and
  associated audio for digital storage media at up to about 1.5 Mbit/s}.
\newblock {International Organization for Standardization}, 1999.

\bibitem{iwata98coarse}
E.~Iwata and K.~Olukotun.
\newblock {Exploiting coarse-grain parallelism in the MPEG-2 algorithm}.
\newblock Technical Report {CSL-TR-98-771}, Stanford University, 1998.

\bibitem{imagine03ieee}
U.~J. Kapasi, S.~Rixner, W.~J. Dally, B.~Khailany, J.~H. Ahn, P.~Mattson, and
  J.~D. Owens.
\newblock Programmable stream processors.
\newblock {\em IEEE Computer}, 2003.

\bibitem{ko05dgt}
D.-I. Ko and S.~S. Bhattacharyya.
\newblock {Dynamic Configuration of Dataflow Graph Topology for DSP System
  Design}.
\newblock In {\em {ICASSP}}, 2005.

\bibitem{bhatta05block}
D.-I. Ko and S.~S. Bhattacharyya.
\newblock {Modeling of Block-Based DSP Systems}.
\newblock {\em {Journal of VLSI Signal Processing}}, 40(3), 2005.

\bibitem{cossap}
J.~Kunkel.
\newblock {COSSAP: A stream driven simulator}.
\newblock In {\em {Int. Workshop on Microelectronics in Communications}}, 1991.

\bibitem{lamb03pldi}
A.~A. Lamb, W.~Thies, and S.~Amarasinghe.
\newblock {Linear Analysis and Optimization of Stream Programs}.
\newblock In {\em {PLDI}}, 2003.

\bibitem{grape-ii}
R.~Lauwereins, M.~Engels, M.~Ad\'e, and J.~Peperstraete.
\newblock {Grape-II: A System-Level Prototyping Environment for DSP
  Applications}.
\newblock {\em {IEEE Computer}}, 28(2), 1995.

\bibitem{lee87static}
E.~Lee and D.~Messershmitt.
\newblock {Static Scheduling of Synchronous Data Flow Programs for Digital
  Signal Processing}.
\newblock {\em IEEE Trans. on Computers}, C-36(1), 1987.

\bibitem{ptolemy03overview}
E.~A. Lee.
\newblock {Overview of the Ptolemy Project}.
\newblock Technical report, UCB/ERL M03/25, UC Berkeley, 2003.

\bibitem{li05alpbench}
M.-L. Li, R.~Sasanka, S.~V. Adve, Y.-K. Chen, and E.~Debes.
\newblock {The ALPBench Benchmark Suite for Complex Multimedia Applications}.
\newblock In {\em {IEEE Int. Symp. on Workload Characterization}}, 2005.

\bibitem{cg03}
W.~R. Mark, R.~S. Glanville, K.~Akeley, and M.~J. Kilgard.
\newblock {Cg: A System for Programming Graphics Hardware in a C-like
  Language}.
\newblock In {\em SIGGRAPH}, 2003.

\bibitem{yelick04msp}
M.~Narayanan and K.~A. Yelick.
\newblock Generating permutation instructions from a high-level description.
\newblock In {\em Workshop on Media and Streaming Processors}, 2004.

\bibitem{neuendorffer04hierarchical}
S.~Neuendorffer and E.~Lee.
\newblock {Hierarchical Reconfiguration of Dataflow Models}.
\newblock In {\em {Conference on Formal Methods and Models for Codesign}},
  2004.

\bibitem{park99spdf2}
C.~Park, J.~Chung, and S.~Ha.
\newblock {Efficient Dataflow Representation of MPEG-1 Audio (Layer III)
  Decoder Algorithm with Controlled Global States}.
\newblock In {\em {IEEE Workshop on Signal Processing Systems}}, 1999.

\bibitem{park02spdf3}
C.~Park, J.~Jung, and S.~Ha.
\newblock {Extended Synchronous Dataflow for Efficient DSP System Prototyping}.
\newblock {\em {Design Automation for Embedded Systems}}, 6(3), 2002.

\bibitem{pazos04soc}
N.~Pazos, P.~Ienne, Y.~Leblebici, and A.~Maxiaguine.
\newblock {Parallel Modelling Paradigm in Multimedia Applications: Mapping and
  Scheduling onto a Multi-Processor System-on-Chip Platform}.
\newblock In {\em {Int. Global Signal Processing Conference}}, 2004.

\bibitem{pelayo01rosa}
F.~L. Pelayo, F.~Cuartero, V.~Valero, D.~Cazorla, and T.~Olivares.
\newblock {Specification and Performance of the MPEG-2 Video Encoder by Using
  the Stochastic Process Algebra: ROSA}.
\newblock In {\em {17th UK Performance Evaluation Workshop}}, 2001.

\bibitem{sermulins05lctes}
J.~Sermulins, W.~Thies, R.~Rabbah, and S.~Amarasinghe.
\newblock {Cache Aware Optimization of Stream Programs}.
\newblock In {\em {LCTES}}, 2005.

\bibitem{shen94overview}
K.~Shen, G.~Cook, L.~Jamieson, and E.~Delp.
\newblock {Overview of parallel processing approaches to image and video
  compression}.
\newblock In {\em {SPIE Conference on Image and Video Compression}}, 1994.

\bibitem{survey97}
R.~Stephens.
\newblock {A Survey of Stream Processing}.
\newblock {\em Acta Informatica}, 34(7), 1997.

\bibitem{streamitcc}
W.~Thies, M.~Karczmarek, and S.~Amarasinghe.
\newblock {StreamIt: A Language for Streaming Applications}.
\newblock In {\em {Int. Conf. on Compiler Construction}}, {2002}.

\bibitem{thies05ppopp}
W.~Thies, M.~Karczmarek, J.~Sermulins, R.~Rabbah, and S.~Amarasinghe.
\newblock Teleport messaging for distributed stream programs.
\newblock In {\em PPoPP}, 2005.

\bibitem{valero02petri}
V.~Valero, F.~L. Pelayo, F.~Cuartero, and D.~Cazorla.
\newblock {Specification and Analysis of the MPEG-2 Video Encoder with
  Timed-Arc Petri Nets}.
\newblock {\em {Electronic Notes in Theoretical Computer Science}}, 66(2),
  2002.

\bibitem{reference-mpeg-c}
{VMPEG (Reference C Code). ftp://ftp.mpegtv}.
\newblock com/pub/mpeg/mssg/mpeg2vidcodec\_v12.tar.gz.

\end{thebibliography}

\end{document}
