% Michal Karczmarek


%\documentclass[times,11pt,twoside,oneandahalfspace]{mitthesis}
%\documentclass[times,11pt,twoside,draft,oneandahalfspace]{mitthesis}
%\documentclass[10pt,oneandahalfspace,twoside,openany]{mitthesis} %\pagestyle{drafthead}
%\documentclass[10pt,oneandahalfspace,twoside,openany,draft]{mitthesis} \pagestyle{drafthead}
%\documentclass[10pt,singlespace,twoside,openany,largemargin]{mitthesis} \pagestyle{drafthead}
%\pagestyle{drafthead}

\documentclass[runningheads,roman]{article}

%\usepackage{lgrind}
\usepackage{verbatim}
\usepackage{epsfig, graphicx}
\usepackage{enumerate}
\usepackage{fullpage}
%\usepackage{doublespace}
%\setstretch{1.05}
%\usepackage{algorithm}
%\usepackage{algorithmic}

%\pagestyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{proof}{Proof}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\newtheorem{definition}{Definition}


\pagestyle{headings}

%\mainmatter

\title{Phased Scheduling of Stream Programs}
\author{Michal Karczmarek, William Thies and Saman
Amarasinghe \\
Laboratory for Computer Science \\
        Massachusetts Institute of Technology \\
        Cambridge, MA  02139 \\
\texttt{\{karczma, thies, saman\}@lcs.mit.edu} \vspace{-24pt}}
\date{}

\begin{document}
\maketitle

\input{commands}

\vspace{0.1in}

\begin{abstract}
Applications that are structured around some notion of a "stream"
are becoming increasingly important and widespread. \cite{Rix98}
provides evidence that streaming media applications are already
consuming most of the cycles on consumer machines, and their use
is continuing to grow. The streaming computation model is
pervasive and ranges from small, embedded systems (ex. cell
phones) to large, computationally powerful machines (ex. cell base
stations).
%
In this paper, we describe a novel technique for scheduling
execution of streaming applications exhibiting hierarchical
properties, such as programs written in {\StreamIt}. {\StreamIt}
programs consist of a number of Streams, which are single-input,
single-output constructs. Streams are constructed from smaller
Streams using hierarchical components, called {\pipeline},
{\splitjoin} and {\feedbackloop}, all of which are Streams themselves.
The smallest computational element of a {\StreamIt} program is a
{\filter}. {\StreamIt} uses a static rate computational model, meaning
that every {\filters} rate of consumption and production of data is
known at compile time.
%
The technique presented here focuses on producing schedules that
are optimized for the amount of space required for buffering and
storing the schedule. A wide variety of real-life applications and
a few synthetic applications are
surveyed. Applications benefit from an average 14.5\% decrease in buffer
requirements, with a peak of 93\% savings in buffer size. No application
requires more space than the most popular technique used today
(Single Appearance Scheduling).
\end{abstract}

%\input{cover}
%\input{contents}
\input{intro}
%\input{streamit}
\input{streamit-asplos}
\input{basics}
\input{hierarchical}
\input{phased}
%\input{constrained}
\input{results}
\input{related}
\input{conclusion}
%\input{pseudocode}

\appendix
\input{appeqs}

%\input{appfigs}

\bibliographystyle{plain}
\bibliography{references}

\end{document}
