With the increasing miniaturization of transistors, wire delays are
becoming a dominant factor in microprocessor performance.  To address
this issue, a number of emerging architectures contain replicated
processing units with software-exposed communication between one unit
and another (e.g., Raw, iWarp, SmartMemories).  However, for
their use to be widespread, it will be necessary to develop compiler
technology that enables a portable, high-level language to execute
efficiently across a range of wire-exposed architectures.

In this paper, we describe our compiler for StreamIt: a high-level,
architecture-independent language for streaming applications.  Our
compiler targets the Raw processor, and we present results using the
cycle-accurate Raw simulator.  Though StreamIt exposes the parallelism
and communication patterns of stream programs, much analysis is needed
to adapt a stream program to a parallel stream processor.  We describe
fission and fusion transformations that can be used to adjust the
granularity of a stream graph, a layout algorithm for mapping a stream
graph to a given network topology, and a scheduling algorithm for
generating a fine-grained static communication pattern for each
computational element.  We consider this work to be a first step
towards a portable programming model for communication-exposed
architectures.

We have a fully functioning compiler, that parallelize StreamIt
applications for the Raw, and a host of optimizations implemented. We
will show that our optimizations are able to increase the performance
upto 300\% and the compiled code was abile to sustan 500 MFLOPS on a
250 MHz Raw processor.
