\section{Conclusion}
\label{sec:conclusion}

In this paper, we describe the StreamIt compiler for the Raw
architecture.  The stream graph of a StreamIt program exposes the data
communication pattern to the compiler while the lack of global
synchronization frees the compiler to radically reoganize the program
for efficient execution on the underline architecture. The StreamIt
compiler demonstrates the power of this flexibility by totally
reoganizing large programs for better load balance. We were able to
map many of programs on to the Raw processor and obtain extremely good
performance, which can go as high as xxx FLOPS.

We introduce a collection of optimizations, vertical and horizontal
filter fusion, vertical and horizontal filter fission and filter
reordering transformations, that are used to radically restructure the
programs.  We show that applying these transformaitons we can map a
high-level stream program, written to reflect the composition of the
application, onto Raw and achieve good processor utilization and load
balance, leading to a factor of three speedup on two applications.

We show how to automatically map a stream graph on to Raw tiles and
produce a communication sschedule and generate efficient code.

Unlike all previous streaming languages, the structured streams of
StreamIt makes it possible for us to approach the optimization and
parallelization problems in a very systermatic manner. It enables us
to define multiple optimizations -- targetting different constructs
and requirements -- compose them and apply them in a hirearchical
manner.

The ability to do global transformations across multiple filters, that
may have originated from very different parts of the application,
makes it possible for the compiler to find optimization opportunities
that may ellude even an experience programmer.  Such capabilities
enables the programmers to write protable streaming applications and
map them efficiently onto any given architecture. This has the
potential of creating a programming standard for the emerging
communication exposed architectures. The StreamIt compiler takes a
fist step towareds this goal.

