{\large ITERATION KEYWORD}

The stream programming paradigm allows the burden of induction variable state to be shifted from the programmer to the compiler.  With the introduction of a new language construct, hereafter referred to as {\tt iter()}, the programmer can take measures to write their programs with data parallelism in mind.  Use of {\tt iter()} potentially provides further benefits in programmability and code readability as well.

As described before, induction variable state requires the user to maintain mutable state, updating such state within the filter invocation, and potentially resetting them when required.  In providing the {\tt iter()} keyword, the programmer can eliminate much of this code simply by arithmetically manipulating {\tt iter()} usage.  

[examples for single induction variable elimination, multiple nested induction variable elimination, varying starting values]

The transformation from filters using induction variable state to using {\tt iter()} is fairly simple for users to implement.  Many of these transformations follow the same approach as induction variable elimination.  The induction variable can be expressed as an arithmetic manipulation of {\tt iter()}.

In having the user define their induction variables using {\tt iter()}, it is possible to eliminate code required to reset or update the induction value.  For large filters, the maintenance of this state may be difficult to follow.  The approach illustrated in the examples shows how it is possible to remove much of the code required to maintain this state with the {\tt iter()} keyword.  

For filters that define multiple induction variables, it is possible to redefine these values in terms of {\tt iter()} as well.  It is not necessary to maintain multiple separate values for state.  Independent and nested induction variable state alike can be defined with use of {\tt iter()}.

This approach of using {\tt iter} to define induction variable state helps users to write filters in a stateless manner.  The user can think about programming their filters in a stateless manner, and thereby exposing data parallelism. 


