\begin{abstract}
With the increasing miniaturization of transistors, wire delays are
becoming a dominant factor in microprocessor performance.  To address
this issue, a number of emerging architectures contain replicated
processing units with software-exposed communication between one unit
and another (e.g., Raw, SmartMemories, TRIPS).  However, for their use
to be widespread, it will be necessary to develop compiler technology
that enables a portable, high-level language to execute efficiently
across a range of wire-exposed architectures.

In this paper, we describe our compiler for StreamIt: a high-level,
architecture-independent language for streaming applications.  We
focus on our backend for the Raw processor.  Though StreamIt exposes
the parallelism and communication patterns of stream programs, some
analysis is needed to adapt a stream program to a software-exposed
processor.  We describe a partitioning algorithm that employs fission
and fusion transformations to adjust the granularity of a stream
graph, a layout algorithm that maps a stream graph to a given network
topology, and a scheduling strategy that generates a fine-grained
static communication pattern for each computational element.

We have implemented a fully functional compiler that parallelizes
StreamIt applications for Raw, including several load-balancing
optimizations.  Using the cycle-accurate Raw simulator, we demonstrate
that these optimizations can improve performance by up to 145\%.  We
consider this work to be a first step towards a portable programming
model for communication-exposed architectures.

\end{abstract}