\section{Applications of SDEP}
\label{sec:others-apps}

This paper applies $\sdep$ to define the semantics of a new language
construct. We believe that there could be additional applications for
$\sdep$ in programming language design and implementation.  Four
potential applications are summarized below.

\subsection{Specifying Latency Constraints}

The latency range included in teleport messaging can be directly
employed as a declarative way to specify latency constraints between
actors.  For instance, our infrastructure can interpret a directive
such as $\mt{maxLatency}(A, B, k)$ to indicate a maximum latency of
$k$ between actors $A$ and $B$.  To implement this directive, the
schedule is constrained as if there were an upstream message from B to
A with latency range $[0,k]$.  This kind of data-centric latency
constraint could be important for reactive applications that need to
produce an output before consuming too many items from the input.

\subsection{Debugging}

An immediate application of $\sdep$ is as part of a graphical
debugging environment for stream programs.  For example, if a user is
stepping through the execution of the work function of actor $B$, it
might become apparent that the errant behavior of $B$ is due to
aspects of the input items that originated in actor $A$.  By utilizing
$\sdep$ information, the debugger can provide the iteration of $A$ in
which the items originated, and the user can continue debugging at
that location.

\subsection{Software-Based Speculation}

Software-based speculation is one approach to improving the
performance of irregular scientific applications~\cite{frank-thesis}.
While the graph-level control flow in Synchronous Dataflow is known at
compile time, there could be unpredictable control flow within the
work function of each actor, and a compiler could attempt to improve
performance by speculatively executing a given path.  However, if the
prediction fails, the results of the speculation might pass outside
the boundaries of the actor and on to other actors in the graph.  In
this case, $\sdep$ provides an exact count of how many iterations the
downstream actors should roll back in order to arrive at the original
state.

\subsection{Program Transformation}

In the realm of scientific computing, a precise notion of dependences
proved essential for developing a robust suite of program analyses and
optimizations.  Representations such as dependence levels~\cite{AK82},
direction vectors~\cite{wolfe82}, and dependence
polyhedra~\cite{Irig88} were important abstractions because they
provided an efficient way to test the validity of program
transformations.  In some ways, $\sdep$ can serve a similar role
within the class of streaming applications.
